\chapter*{Introduction}
\section*{Syllabus}
The first part of the course deals with:
\begin{itemize}
    \item Algebraic Tori, their actions and representations
    \item Affine toric varieties (with monoids) $\leftrightarrow$ cones in some $\R^n$
    \item Projective toric varieties $\leftrightarrow$ polytopes in some $\R^n$
    \item General toric varieties $\leftrightarrow$ fans in $\R^n$
\end{itemize}
We will then deal with (suject to change)
\begin{itemize}
    \item Divisors/line bundles on toric varieties
    \item Cox ring of a toric variety
    \item Cohomology of divisors
    \item Toric morphisms and resolution of singularities
    \item and more...?
\end{itemize}
The main reference for this course \textemdash ``Toric varieties" by Cox, Little, Schenck \cite{cox2011toric} \textemdash is available in the same folder as this PDF.

\section*{What is the course about?}
We will work over an algebraically closed field (and we will be lax about the characteristic of the field). In \cite{cox2011toric} the authors work over $\C$ but many results hold more generally.
\medskip

The main goal of the course is understanding toric varieties:
\begin{definition}[Toric variety]
An $n$-dimensional toric variety $X$ is a (normal) $k$-variety equipped with an open immersion of an $n$-dimensional torus $T\subseteq X$, where $T\cong (k^\ast)^n$, and an action $T\times T\to T$ which extends to the whole of $X$ \footnote{that is, it extends to a $T\times X\to X$}.
\end{definition}

\begin{remark}
Normality is a standard assumption that we'll make at some point but some things work without it.
\end{remark}

We'll see that the geometry of such an object is encoded in a combinatorial gadget, converting problems in algebraic geometry to problems in combinatorics, which is sometimes convenient.

The opposite reduction is also possible and has been used historically. One of the main examples of a combinatorial problem being solved via the geometry of toric varieties is
\begin{example}[McMullen's ``$g$-conjecture"]

The then conjecture, and now theorem, is a characterization of the $f$-vectors of simple polytopes\footnote{for now a simple polytope is the convex hull of a finite subset of $\R^n$}.

\begin{definition}[$f$-vectors]
If $P$ is a polytope, its \textbf{$f$-vector} is 
\[(f_0(P),\cdots, f_d(P)),\quad \text{where $d=\dim P$}\]
and $f_i(P)$ is the number of $i$-dimensional faces. We may set $f_{-1}(P)=1$.
\end{definition}

It's reasonable to ask ourselves which $f$-vectors can appear. We may define the \textbf{$h$-vector} by setting
\[\sum_{i=0}^d f_i (t-1)^i=\sum_{i=0}^dh_i t^i,\quad \text{i.e. }h_i=\sum_{j=i}^d(-1)^{j-i}\binom{j}if_j,\ h_{-1}=0.\]
It was a known theorem that the $h$-vector of a simple polytope is palindromic (i.e. $h_i=h_{d-i}$).
From the $h$-vector we obtain the \textbf{$g$-vector} by setting $g_i=h_i-h_{i-1}$. 

The conjecture was that
\begin{theorem}[$g$-conjecture]
$f=(f_0,\cdots, f_d)\in \N^{d+1}$ is the $f$-vector of a simple polytope if
\begin{enumerate}
\item $h_i=h_{d-1}$ for all $0\leq i\leq \floor{d/2}$
\item $g_i\geq 0$ for all $0\leq i\leq \floor{d/2}$
\item $(g_1,\cdots, g_{\floor{d/2}})$ is a ``Macauly vector" if, when we write (uniquely)
\[g_i=\binom{n_i}i+\cdots+\binom{n_{r_i}}{r_i}\]
with $n_i>n_{i-1}>\cdots>n_{r_i}$ then
\[g_{i+1}\leq \binom{n_i+1}{i+1}+\cdots+\binom{n_{r_i}+1}{r_i+1}\]
\end{enumerate}
\end{theorem}
Stanley proved necessity using toric varieties. He proved that the $g$-vector of a simple polytope is the vector of dimensions for some cohomology ring of the associated toric variety.


Later McMullen found a completely combinatorial proof but for some time the only proof of this combinatorial fact passed through the geometry of toric varieties.
\end{example}